KIN: A method to infer relatedness from low-coverage ancient DNA

### 2022-05-18 paper_overleaf/main.tex
Updates from Overleaf

Identifying related individuals is a common task in genetic studies. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity \cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}.

@@

However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from an out-of-sample reference panel to calculate the probability of IBD given IBS. There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022}.

@@

One limitation of our approach is that it assumes a single population. In case of a highly structured population, KIN may show inaccurate inference of $p_0$ causing inaccurate relatedness inference. Also, our method makes the assumption, that the median pairwise genetic difference in the population reflects the population diversity $p_0$, which fails if almost all individuals in the dataset are related. The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.

@@

\section{Competing interests}
The authors declare that they have no competing interests.

>>>>

Identifying related individuals is a common task in genetic studies. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between an individual applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity \cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}.

@@

However, since it is not possible to directly observe IBD segments, a common approach is to first identify segments of the genome that are Identical by State (IBS) and to use population allele frequencies obtained from an out-of-sample reference panel to calculate the probability of IBD given IBS. There are several methods that incproporate reference panel allele frequencies, phase information, recombination maps, or genotype calls to co-estimate IBD and the relatedness coefficient \cite{huff_maximum-likelihood_2011,li_relationship_2014,thornton_estimating_2012, boehnke_accurate_1997,lynch_estimation_1999, purcell_plink_2007,manichaikul_robust_2010,gusev_whole_2009,nyerki_optimized_2022,browning_fast_2011,li_accurate_2014}.

@@

One limitation of our approach is that it assumes a single population. In case of a highly structured population, KIN may show inaccurate inference of $p_0$ causing inaccurate relatedness inference. Also, our method makes the assumption, that the median pairwise genetic difference in the population reflects the population diversity $p_0$, which fails if almost all individuals in the dataset are related. We may get around this problem by using an estimate of $p_0$, calculated from known a pair of unrelated individuals from same population, or another population with similar diversity. The current implementation of KIN is restricted to six relatedness cases we expect to be most common, but it might be feasible to extend it to other cases, such as double first cousins, using a corresponding IBD state transition matrix.

@@

\section{Data and material availability}
An open-source implementation of KIN and KINgaroo in python along with a toy example dataset, and the scripts to generate our test simulations are available on GitHub \url{https://github.com/DivyaratanPopli/Kinship_Inference}. We have deposited the version of software used in the manuscript on Zenodo (). The analysed datasets from Bronze Age Lech Valley were generated in a previous study \cite{mittnik_kinship-based_2019}. Chagyrskaya and Okladnikov dataset were generated from a study currently in press \cite{laurits_skov_genetic_nodate}, and the data will be uploaded to European Nucleotide Archive upon publication.

\section{Ethics declarations}
\subsection{Ethics approval and consent to participate}
Not applicable.

\subsection{Consent for publication}
Not applicable.

\subsection{Competing interests}
The authors declare that they have no competing interests.
